Starcode: sequence clustering based on all-pairs search

### 2014-11-03 doc/document.tex
Updated benchmark figure. Text of the manuscript in progress.

\begin{methods}
\section{Methods}
\subsection{Inexact string matching using tries}
The matching method of starcode is based on a variation of the
Needleman-Wunsch (NW) algorithm \citep{pmid5420325}. In the
original algorithm (Figure~\ref{fig:NW}a), the Levenshtein
distance between two sequences is found by applying a recurrence
relation throughout a matrix of $mn$ terms (the edit matrix),
where $m$ and $n$ are the respective sequence lengths. The complexity
of this dynamic programming approach is $O(mn)$.

@@

In many instances, the information of interest is to find out
whether the sequences are $\tau$-matches (\textit{i.e.} their distance
is less then or equal to a fixed threshold $\tau$).
In that case, the complexity can be reduced to $O(\tau \min(m,n))$.
Instead of computing all the terms of the edit matrix, it is initialized
as shown on Figure~\ref{fig:NW}b and only the terms around the diagonal
are computed. If a diagonal term has a value greater than $\tau$,
the process is halted because the sequences are not $\tau$-matches.

This method can be used to match sequences against a prefix tree,
also known as a trie \citep{ukkonen}. The terms of the edit matrix
are updated row-wise while a depth-first search traverses the trie
(Figure~\ref{fig:trie}). Every time a node is visited,
a row is computed, and every time the search backtracks, a row is
erased. If the threshold value $\tau$
is exceeded for a diagonal term, the Levenshtein distance for all the
downstream sequences is also necessarily greater than $\tau$. Therefore,
no more hits are to be discovered in this path and the depth-first
search backtracks to the parent node. When the process halts, every
tail node (corresponding to a sequence of the database) on the path
of this search is a $\tau$-match of the query. This method is
efficient because it eliminates large areas of the search space, and
because the NW comparison of the query with each prefix of the database
is computed only once.

\begin{figure}[!tpb]
\centerline{\includegraphics{trie.pdf}}
\caption{NW algorithm on tries. Each sequence of the index is a
path in the trie. The query GTTGCA is written at the top of the matrix,
which is initialized as shown on Figure~\ref{fig:NW}b. The trie is
traversed by a depth-first search (red path). At each
depth, the node added to the path is written on the left of the edit
matrix and the row is computed. Checkpoints from 1 to 4 (circled red
numbers) show the state of the edit matrix as the search proceeds. The
node labeled 3 is a leaf and thus corresponds to a 2-match of the
query. After discovering the hit, the search path backtracks to the
node labeled 2 and the last rows of the edit matrix are erased.
The search path then goes to the node labeled 4, in which case the
newly computed diagonal cell exceeds the threshold (circled in red).
Even if this node has children, they are not visited (red crosses)
because there is no 2-match to discover.}
\label{fig:trie}
\end{figure}

\subsection{The poucet search algorithm}
The search strategy can be further improved. If two consecutive
queries share a prefix of length $k$, the succession of computations
up to the $k$-th row of the edit matrix will be exactly the same
for both queries. Therefore, computation intermediates can be
stored in the nodes of the trie, so that the next trie search can start
at depth $k$. However, storing the rows of the edit matrix in the nodes
meets some difficulty. Indeed, on the $k$-th row, the terms on the
right side of the diagonal depend on characters that are not shared
between the two queries. This issue is solved by storing in each
node a combination of row and column terms that form an angle shape,
looking like a horizontally flipped L (Figure~\ref{fig:poucet}).
Using this structure, the computation intermediates stored in a
node at depth $k$ depend only on the first $k$ characters of the
query.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics{poucet.pdf}}
\caption{Poucet search algorithm. The algorithm proceeds with the
same principles as shown on Figure~\ref{fig:trie} with the difference
that the edit matrix is not updated row-wise, but along a horizontally
flipped L. As the depth-first search proceeds, these values
are stored in the nodes of the trie. Since the values in the vertical
part of the flipped L are the same for every child of a node,
they are computed only once (purple arrow). The values in the grey
cells will be computed as the search path (red) visits the node.
Storing the intermediates in the nodes allows the next query to
restart at depth $k$ if it shares a common prefix of length
$k$ with the current query.}
\label{fig:poucet}
\end{figure}

@@

Before the search, segments of the query are looked up in
the indexes, and in case no match is found, the trie search is
omitted because this query has no $\tau$-match. If at least one
segment is found, the trie search is performed. As mentioned above,
segments shared between the query and a $\tau$-match can be shifted
up to $\tau$ nucleotides. For this reason, shifted segments are
looked up in the indexes according to the following scheme ensuring
that no match can be missed.
(Figure~\ref{fig:filtration}). The rightmost segment is looked up
in the $\tau$+1-th index, the second rightmost segment and the
contiguous segments shifted by 1 nucleotides are are looked up in the
$\tau$-th index and so on until the first segment and its
contiguous segments shifted by up to $\tau$ nucleotides are
looked up in the first index.

@@

\subsection{Clustering}
The default clustering algorithm of starcode is designed to correct
sequencing error. This method uses message passing \citep{mackay}
to identfy and count ``canonical'' sequences (also referred to as
centroids in the clustering terminology). Each sequence transfers its
read count to its closest $\tau$-match provided the latter has at least 5
times more counts. If the condition is not met, the transfer does
not take place. If the sequence has several equally close
$\tau$-matches, the counts are split equally among them. The
process is repeated recursively, starting from sequences with
lowest read count. The sequences with a positive read count at
the end of the process are considered canonical. Clusters consist
of all the sequences transferring their read counts to the same
canonical sequence (sequence transferring their read counts to
different canonicals are discarded).

@@

\subsection{Basic performance}
\label{performance}
We measured basic performance and scalability metrics of starcode on a
dataset of pseudo random sequences (Figure~\ref{fig:perf}). The standard
configuration consists of a set of 1,000,000 sequences of length 40 running
on 1 thread and with a maximum Levenshtein distance of 3. To test the
scalability as a function of a single parameter, only the parameter under
study was modified whereas the others were kept constant.

Figure~\ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic scale
the trend is a straight line with slope 1.5, suggesting that the
running time complexity of starcode is lower than quadratic (unlike
the naive implementation of all-pairs search).
Figure~\ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
The reason is that the trie fans out exponentially and the search bails
out at a greater depth as the maximum distance increases.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure~\ref{fig:perf}c). Beyond a
threshold length, the filtering algorithm starts to be efficient, and
most of the queries are resolved without searching the trie.
Finally, we show the scalability of starcode with increasing
number of threads in Figure~\ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 12 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is that there is insufficient memory bandwidth to satisfy
the increased demand of memory accesses due to hardware limitations.

@@

\subsection{Benchmark}
We benchmarked starcode against the sequence clustering algorithms
slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
slidesort is an all-pairs search algorithm, it was included in the
benchmark because sequence comparison is the most computationally
intensive step of the sequence clustering problem. Rainbow runs
exclusively on paired-end reads, while the other tools run on single
reads.

%% Synthetic data %%
The performance of sequence clustering algorithms can be sensitive
to the size of the clusters in the dataset, which in many applications
is not known \textit{a priori}. We therefore set up a benchmark on
artificial datasets to test the accuracy and the scaling of the
tools on a known cluster structure. We generated 4 datasets of 1 million
50-mers arranged in 1 to 1,000 clusters. Each cluster consisted
of 100 repeats of the same centroid sequence, plus satellites
derived from the centroid by incorporating 3 errors including at
most 1 indel. The number of satellites per cluster ranged from
999900 to 900. Evaluating the exactness of slidesort on these datasets
was problematic because the number of 3-matches in each dataset is
not known (pairs of statellites in the same cluster may be 3-matches
or not). For this reason we only compared the number of pairs found
by starcode with the number of pairs found by slidesort. The 
outcome of the test is summarized in
Figure~\ref{fig:benchmark}.

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets. We also observed that slidesort found 5-10\% less
3-matches than starcode on all the datasets. We were surprised by
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons is feasible
(the starcode repository contains such a dataset along with the
instructions to reproduce our observations, see
http://github.com/gui11aume/starcode/misc). In conclusion, starcode
was the only tool to achieve perfect precision.

@@

%% Biological datasets %%

We benchmarked sequence clustering algorithms on the problem of
clustering TRIP barcodes \citep{pmid23953119}. Briefly, the principle
of TRIP (Thousands of Reporters Integrated in Parallel) is to tag
reporter transcripts with random barcodes and measure the abundance
of barcodes in the RNA as a proxy for gene expression. There is no
reference to match aberrant barcodes against, because the tagging
sequences are unknown.

@@

\subsection{Identifying enriched sequence motifs}
Sequence motifs are thought to play an important role in DNA metabolism.
Key regulators, such as transcription factors, nucleosomes and non
coding RNAs have sequence preferences targeting them to the sites where
they act. Identifying those sequences is a way to pinpoint the
regulators and the mechanisms they are involved in. However, the
sequence motifs are not strictly identical at different sites, hence
they are better identified by inexact matching. This problem becomes
computationally difficult for long motifs (above 12-13 nucleotides)
because of the combinatorial scaling. But as motifs become longer, the
problem of identifying abundant inexact matches becomes similar to
barcode clustering. We reasoned that starcode could also be used for
the task of identifying biologically meaningful sequence motifs.

@@

The current version of starcode has been primarily optimized for speed.
The memory footprint depends on the number of sequences to cluster
(because the sequences of the input set are loaded in memory as a set of
of tries) and on the mean number of matches per sequence. Every match is
stored in memory until the clustering phase, which may represent a large
overhead if the dataset is dense. As counterintuitive as it may seem,
long queries will usually impose a lower memory footprint because the
matches between the sequences are more sparse. We have shown that the
running time will also be shorter thanks to the lookup table
search (Figure~\ref{fig:perf}c).

Perhaps the most surprising element of this study is that an exact
algorithm can compete with extremely fast heuristics. This will no
longer be true when clustering divergent sequences because the Levenshtein
distance will have to be increased, leading to exponentially longer
running times (Figure~\ref{fig:perf}b). However, for the imporant
practical case that the divergence is driven by sequencing errors,
starcode illustrates that there is still room for algorithmic innovations
that can outperform heuristics. The idea of the poucet search seems
simple in retrospect, yet it is a powerful way to tap into the data
structuration provided by string sorting. This principle could find some
applications in algorithms used in various fields.

>>>>

\begin{methods}
\section{Methods}
\subsection{Inexact string matching using tries}
The matching method of starcode is based on a variation of the
Needleman-Wunsch (NW) algorithm \citep{pmid5420325}. In the
original algorithm (Figure \ref{fig:NW}a), the Levenshtein
distance between two sequences is found by applying a recurrence
relation throughout a matrix of $mn$ terms (the edit matrix),
where $m$ and $n$ are the respective sequence lengths. The complexity
of this dynamic programming approach is $O(mn)$.

@@

In many instances, the information of interest is to find out
whether the sequences are $\tau$-matches (\textit{i.e.} their distance
is less then or equal to a fixed threshold $\tau$).
In that case, the complexity can be reduced to $O(\tau \min(m,n))$.
Instead of computing all the terms of the edit matrix, it is initialized
as shown on Figure \ref{fig:NW}b and only the terms around the diagonal
are computed. If a diagonal term has a value greater than $\tau$,
the process is halted because the sequences are not $\tau$-matches.

This method can be used to match sequences against a prefix tree,
also known as a trie \citep{ukkonen}. The terms of the edit matrix
are updated row-wise while a depth-first search traverses the trie
(Figure \ref{fig:trie}). Every time a node is visited,
a row is computed, and every time the search backtracks, a row is
erased. If the threshold value $\tau$
is exceeded for a diagonal term, the Levenshtein distance for all the
downstream sequences is also necessarily greater than $\tau$. Therefore,
no more hits are to be discovered in this path and the depth-first
search backtracks to the parent node. When the process halts, every
tail node (corresponding to a sequence of the database) on the path
of this search is a $\tau$-match of the query. This method is
efficient because it eliminates large areas of the search space, and
because the NW comparison of the query with each prefix of the database
is computed only once.

\begin{figure}[!tpb]
\centerline{\includegraphics{trie.pdf}}
\caption{NW algorithm on tries. Each sequence of the index is a
path in the trie. The query GTTGCA is written at the top of the matrix,
which is initialized as shown on Figure \ref{fig:NW}b. The trie is
traversed by a depth-first search (red path). At each
depth, the node added to the path is written on the left of the edit
matrix and the row is computed. Checkpoints from 1 to 4 (circled red
numbers) show the state of the edit matrix as the search proceeds. The
node labeled 3 is a leaf and thus corresponds to a 2-match of the
query. After discovering the hit, the search path backtracks to the
node labeled 2 and the last rows of the edit matrix are erased.
The search path then goes to the node labeled 4, in which case the
newly computed diagonal cell exceeds the threshold (circled in red).
Even if this node has children, they are not visited (red crosses)
because there is no 2-match to discover.}
\label{fig:trie}
\end{figure}

\subsection{The poucet search algorithm}
The search strategy can be further improved. If two consecutive
queries share a prefix of length $k$, the succession of computations
up to the $k$-th row of the edit matrix will be exactly the same
for both queries. Therefore, computation intermediates can be
stored in the nodes of the trie, so that the next trie search can start
at depth $k$. However, storing the rows of the edit matrix in the nodes
meets some difficulty. Indeed, on the $k$-th row, the terms on the
right side of the diagonal depend on characters that are not shared
between the two queries. This issue is solved by storing in each
node a combination of row and column terms that form an angle shape,
looking like a horizontally flipped L (Figure \ref{fig:poucet}).
Using this structure, the computation intermediates stored in a
node at depth $k$ depend only on the first $k$ characters of the
query.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics{poucet.pdf}}
\caption{Poucet search algorithm. The algorithm proceeds with the
same principles as shown on Figure \ref{fig:trie} with the difference
that the edit matrix is not updated row-wise, but along a horizontally
flipped L. As the depth-first search proceeds, these values
are stored in the nodes of the trie. Since the values in the vertical
part of the flipped L are the same for every child of a node,
they are computed only once (purple arrow). The values in the grey
cells will be computed as the search path (red) visits the node.
Storing the intermediates in the nodes allows the next query to
restart at depth $k$ if it shares a common prefix of length
$k$ with the current query.}
\label{fig:poucet}
\end{figure}

@@

Before the search, segments of the query are looked up in
the indexes, and in case no match is found, the trie search is
omitted because this query has no $\tau$-match. If at least one
segment is found, the trie search is performed. As mentioned above,
segments shared between the query and a $\tau$-match can be shifted
up to $\tau$ nucleotides. For this reason, shifted segments are
looked up in the indexes according to the following scheme ensuring
that no match can be missed.
(Figure \ref{fig:filtration}). The rightmost segment is looked up
in the $\tau$+1-th index, the second rightmost segment and the
contiguous segments shifted by 1 nucleotides are are looked up in the
$\tau$-th index and so on until the first segment and its
contiguous segments shifted by up to $\tau$ nucleotides are
looked up in the first index.

@@

\subsection{Clustering}
The default clustering algorithm of starcode is designed to correct
sequencing error. This method uses message passing \citep{mackay}
to identfy and count ``canonical'' sequences (also referred to as
centroids in the clustering terminology). Each sequence transfers its
read count to its closest $\tau$-match provided the latter has at least 5
times more counts. If the condition is not met, the transfer does
not take place. If the sequence has several equally close
$\tau$-matches, the counts are split equally among them. The
process is repeated recursively, starting from sequences with
lowest read count. The sequences with a positive read count at
the end of the process are considered canonical. Clusters consist
of all the sequences transferring their read counts to the same
canonical sequence (sequence transferring their read counts to
different canonicals are discarded). Note that the radius of the
clusters can be higher than the maximum distance used for matching.

@@

\subsection{Presentation and basic performance}
\label{performance}
Starcode is a general purpose DNA sequence clustering tool with a
strong focus on error correction. Errors are assumed to be mismatches,
insertions or deletions (the implementation presented here matches
sequences with up to 8 errors). The input sequences can be single
or paired-end reads, with an upper limit of 1024 nucleotides (512
for paired-end). Sequences may be of variable length, they may
be trimmed and filtered for quality or not. Input formats may be raw
sequence, fasta or fastq (in which case starcode ignores the quality).
By default, starcode outputs sequence clusters, it can also be used
to filter out redundant sequences from input files. By default,
clustering is performed under the assumption that divergence occurs
from experimental errors (sequencing errors, PCR mutations
\textit{etc}), a more general algorithm is also available for other
clustering problems (an example of which is given in section
\ref{section:motifs}).

We show the basic performance and scalability of starcode on a
dataset of pseudo random sequences (Figure \ref{fig:perf}). The standard
configuration is a set of 1,000,000 sequences of length 40, running
on 1 thread and with a maximum Levenshtein distance of 3. Only one
parameter is modified whereas the others were kept constant.

Figure \ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic
scale the trend is a straight line with slope 1.5, suggesting that the
running time complexity of starcode is lower than quadratic (the
naive implementation of all-pairs search). Note that the sequences of
this dataset have no match, see section \ref{section:benchmark} for
an evaluation of the performance on more realistic datasets.
Figure \ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
The reason is that the trie fans out exponentially and the search bails
out at a greater depth as the maximum distance increases.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure \ref{fig:perf}c). Beyond a
threshold length, the filtering algorithm starts to be efficient, and
most of the queries are resolved without searching the trie.
Finally, we show the scalability of starcode with increasing number
of threads in Figure \ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 12 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is that there is insufficient memory bandwidth to satisfy
the increased demand of memory accesses due to hardware limitations.

@@

\subsection{Benchmark}
\label{section:benchmark}
We benchmarked starcode against the sequence clustering algorithms
slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
slidesort is an all-pairs search algorithm, it was included in the
benchmark because sequence comparison is the most computationally
intensive step of the sequence clustering problem. Rainbow runs
exclusively on paired-end reads, while the other tools run on single
reads.

%% Synthetic data %%
The performance of sequence clustering algorithms can be sensitive
to the size of the clusters in the dataset, which in many applications
is not known \textit{a priori}. We therefore set up a benchmark on
artificial datasets to test the accuracy and the scaling of the
tools on a known cluster structure. We generated 4 datasets of 1 million
50-mers arranged in 1 to 1,000 clusters. Each cluster consisted
of 100 repeats of the same centroid sequence, plus satellites
derived from the centroid by incorporating 3 errors including at
most 1 indel. The number of satellites per cluster ranged from
999900 to 900. Evaluating the exactness of slidesort on these datasets
was problematic because the number of 3-matches in each dataset is
not known (pairs of statellites in the same cluster may be 3-matches
or not). For this reason we only compared the number of pairs found
by starcode with the number of pairs found by slidesort. The 
outcome of the test is summarized in
Figure \ref{fig:benchmark}.

@@

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets. We also observed that slidesort found 5-10\% less
3-matches than starcode on all the datasets. We were surprised by
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons is feasible
(the starcode repository contains such a dataset along with the
instructions to reproduce our observations, see
http://github.com/gui11aume/starcode/misc).

The running time of the different tools as a function of the size
of the clusters is shown on Figure \ref{fig:benchmark}c. For seed
and cd-hit it decreases as the cluster size increases, whereas
for starcode and slidesort it increases. In spite of this trend,
the performance of starcode remains competitive, even for clusters
of 1 million sequences. The memory usage is shown in Figure
\ref{fig:benchmark}c.
In conclusion, starcode was the only tool to achieve perfect precision
on these datasets. It also maintains competitive performance in
running time and memory usage.

@@

%% Biological datasets %%
The performance on artificial data is not always in agreement with
the performance on experimental datasets. In typical experiments,
the sizes of the clusters may be uneven, and the reads may contain
near constant regions that may degrade the performance.
We benchmarked sequence clustering algorithms on the problem of
clustering TRIP barcodes \citep{pmid23953119}. Briefly, the principle
of TRIP (Thousands of Reporters Integrated in Parallel) is to tag
reporter transcripts with random barcodes and measure the abundance
of barcodes in the RNA as a proxy for gene expression. There is no
reference to match aberrant barcodes against, because the tagging
sequences are unknown.

@@

\subsection{Identifying enriched sequence motifs}
\label{section:motifs}
Sequence motifs are thought to play an important role in DNA metabolism.
Key regulators, such as transcription factors, nucleosomes and non
coding RNAs have sequence preferences targeting them to the sites where
they act. Identifying those sequences is a way to pinpoint the
regulators and the mechanisms they are involved in. However, the
sequence motifs are not strictly identical at different sites, hence
they are better identified by inexact matching. This problem becomes
computationally difficult for long motifs (above 12-13 nucleotides)
because of the combinatorial scaling. But as motifs become longer, the
problem of identifying abundant inexact matches becomes similar to
barcode clustering. We reasoned that starcode could also be used for
the task of identifying biologically meaningful sequence motifs.

@@

The current version of starcode has been primarily optimized for speed.
The memory footprint depends on the number of sequences to cluster
(because the sequences of the input set are loaded in memory as a set of
of tries) and on the mean number of matches per sequence. Every match is
stored in memory until the clustering phase, which may represent a large
overhead if the dataset is dense. As counterintuitive as it may seem,
long queries will usually impose a lower memory footprint because the
matches between the sequences are more sparse. We have shown that the
running time will also be shorter thanks to the lookup table
search (Figure \ref{fig:perf}c).

Perhaps the most surprising element of this study is that an exact
algorithm can compete with extremely fast heuristics. This will no
longer be true when clustering divergent sequences because the Levenshtein
distance will have to be increased, leading to exponentially longer
running times (Figure \ref{fig:perf}b). However, for the imporant
practical case that the divergence is driven by sequencing errors,
starcode illustrates that there is still room for algorithmic innovations
that can outperform heuristics. The idea of the poucet search seems
simple in retrospect, yet it is a powerful way to tap into the data
structuration provided by string sorting. This principle could find some
applications in algorithms used in various fields.
